Symbolic Weighted Language Models and Parsing over Infinite Alphabets

### 2021-05-05 main.tex
1 passe def. transducers

\subsection{Definitions}
%(SWT)
\begin{definition}
A \emph{symbolic-weighted transducer} (\SWT) 
$T$ over the input and output alphabet $\Sigma$ and $\Delta$ 
and the semiring $\Semiring$ is a tuple
$T = \< Q, \init, \wei, \final >$,
where $Q$ is a finite set of states, 
$\mathsf{in} : Q \to \Semiring$, 
respectively $\mathsf{out} : Q \to \Semiring,$
are functions defining the weight for entering, 
respectively leaving, a state, 
and $\wei$ is a transition %weight 
function from $Q \times Q$ into %$\Phi$.
$\< \Phi_\epsilon, \Phi_\Sigma, \Phi_\Delta, \Phi_{\Sigma, \Delta}>$.
\end{definition}
%
We extend the above transition function into a function from
$Q \times (\Sigma \cup \{ \epsilon \}) \times (\Delta \cup \{ \epsilon \}) \times Q$
into $\Semiring$, also called $\wei$ for simplicity, such that 
for all $q, q' \in Q$, $a \in \Sigma$, $b \in \Delta$, 
and with 
$\< \phi_\epsilon, \phi_\Sigma, \phi_\Delta, \phi_{\Sigma, \Delta}> = \wei(q, q')$, 
\[
\begin{array}{rcl}
\wei(q, \epsilon, \epsilon, q') & = & \phi_\epsilon\\
\wei(q, a, \epsilon, q') & = & \phi_\Sigma(a)\\
\wei(q, \epsilon, b, q') & = & \phi_\Delta(b)\\
\wei(q, a, b, q') & = & \phi_{\Sigma, \Delta}(a, b)
\end{array}      
\]
%More precisely, $Q \times Q$, 
%resp. $Q \times \Sigma \times Q$,
%$Q \times \Delta \times Q$,
%$Q \times \Sigma \times \Delta \times Q$,
%into resp. $\Phi_0$, $\Phi_\Sigma$, $\Phi_\Delta$ $\Phi_{\Sigma, \Delta}$.

These functions $\phi$ act as guards for the transducer's transitions, 
preventing a transition when they return the absorbing $\zero$ of $\Semiring$.

The symbolic-weighted transducer $T$ defines a mapping 
from the pairs of strings of $\Sigma^* \times \Delta^*$ 
into the weights of~$\Semiring$,
based on the following intermediate function $\weight_T$
defined recursively for every $q, q' \in Q$, 
%$a \in \Sigma$, $b \in \Delta$ 
for every strings of $s \in \Sigma^*$, $t \in \Delta^*$:
\[
\begin{array}{rccl}
\weight_T(q, s, t, q') & = & & \wei(q, \epsilon, \epsilon, q')\\
 & & \oplus \displaystyle\bigoplus_{\begin{array}{c}
                                      \scriptstyle q'' \in Q\\[-2pt]
                                      \scriptstyle s = au, a \in \Sigma
                                      \end{array}} &
     \wei(q, a, \epsilon, q'') \otimes \weight_T(q'', u, t, q')\\
 & & \oplus \displaystyle\bigoplus_{\begin{array}{c}
                                    \scriptstyle q'' \in Q\\[-2pt]
                                    \scriptstyle t = bv, b \in \Delta
                                    \end{array}} &
      \wei(q, \epsilon, b, q'') \otimes \weight_T(q'', s, v, q')\\ 
      & & \oplus \displaystyle\bigoplus_{\begin{array}{c}
                                         \scriptstyle q'' \in Q\\[-2pt]
                                         \scriptstyle s = au, a \in \Sigma\\[-2pt]
                                         \scriptstyle t = bv, b \in \Delta
                                         \end{array}} &
      \wei(q, a, b, q'') \otimes \weight_T(q'', u, v, q')\\ 
\end{array}
\]
Recall that by convention, an empty sum with $\oplus$ is $\zero$.

The above expression of $\weight_T$ can be seen as a stateful definition of 
an edit-distance between a word $s \in \Sigma^*$ and a word $t \in \Delta^*$,
see also~\cite{Mohri03ijfcs}.
Intuitively, 
$\wei(q, a, \epsilon, q'')$, in the second line, is the cost of 
the deletion of a symbol $a \in \Sigma$ in $s$, 
$\wei(q, \epsilon, b, q'')$, in the third line, is the cost 
of the deletion of insertion of  $b \in \Delta$ in $s$, 
$\wei(q, a, b, q'')$, in the last line, is the cost 
of the substitution of  $a \in \Sigma$ by $b \in \Delta$.
The cost of a sequence of such operations transforming $s$ into $t$ is 
the product with $\otimes$ of the costs of individual operations involved,
and the distance between $s$ and $t$ is the sum with $\oplus$
of all such cost products.

>>>>

\subsection{Definitions}
%(SWT)
\begin{definition}
A \emph{symbolic-weighted transducer} (\SWT)~$T$ 
over the input and output alphabets~$\Sigma$ and $\Delta$ 
and the semiring $\Semiring$ is a tuple
$T = \< Q, \init, \vec{\wei}, \final >$,
where $Q$ is a finite set of states, 
$\mathsf{in} : Q \to \Semiring$, 
respectively $\mathsf{out} : Q \to \Semiring,$
are functions defining the weight for entering, 
respectively leaving, a state, 
and $\vec{\wei}$ is a tuplet made of the 4 of transition functions 
$\wei_\epsilon$, $\wei_\Sigma$, $\wei_\Delta$, and $\wei_{\Sigma, \Delta}$
from $Q \times Q$ into respectively
$\Phi_\epsilon$, $\Phi_\Sigma$, $\Phi_\Delta$, and $\Phi_{\Sigma, \Delta}$.
\end{definition}
%
We synthesize the above 4-uplet of transition functions into a unique 
function~$\wei$ from
$Q \times (\Sigma \cup \{ \epsilon \}) \times (\Delta \cup \{ \epsilon \}) \times Q$
into $\Semiring$, %also called $\wei$ for simplicity, 
such that, for all $q, q' \in Q$, $a \in \Sigma$, and $b \in \Delta$, 
%and with 
%$\< \phi_\epsilon, \phi_\Sigma, \phi_\Delta, \phi_{\Sigma, \Delta}> = \wei(q, q')$, 
\[
\begin{array}{rcll}
\wei(q, \epsilon, \epsilon, q') & = &  \wei_\epsilon(q, q'),\\ %\phi_\epsilon\\
\wei(q, a, \epsilon, q') & = & \phi_\Sigma(a) & %\wei_\Sigma(q, q')(a)
\quad\mathrm{where~} \phi_\Sigma = \wei_\Sigma(q, q'),\\
\wei(q, \epsilon, b, q') & = & \phi_\Delta(b) & 
\quad\mathrm{where~} \phi_\Delta = \wei_\Delta(q, q'),\\
\wei(q, a, b, q') & = & \phi_{\Sigma, \Delta}(a, b) & 
\quad\mathrm{where~} \phi_{\Sigma, \Delta} = \wei_{\Sigma, \Delta}(q, q').\\
\end{array}      
\]
%More precisely, $Q \times Q$, 
%resp. $Q \times \Sigma \times Q$,
%$Q \times \Delta \times Q$,
%$Q \times \Sigma \times \Delta \times Q$,
%into resp. $\Phi_0$, $\Phi_\Sigma$, $\Phi_\Delta$ $\Phi_{\Sigma, \Delta}$.

The symbolic-weighted transducer $T$ defines a mapping 
from the pairs of strings of $\Sigma^* \times \Delta^*$ 
into the weights of~$\Semiring$,
based on the following intermediate function $\weight_T$
defined recursively for every $q, q' \in Q$, 
%$a \in \Sigma$, $b \in \Delta$ 
for every strings of $s \in \Sigma^*$, $t \in \Delta^*$:
\[
\begin{array}{rccl}
\weight_T(q, s, t, q') & = & & \wei(q, \epsilon, \epsilon, q')\\
 & & \oplus \displaystyle\bigoplus_{\begin{array}{c}
                                      \scriptstyle q'' \in Q\\[-2pt]
                                      \scriptstyle s = au, a \in \Sigma
                                      \end{array}} &
     \wei(q, a, \epsilon, q'') \otimes \weight_T(q'', u, t, q')\\
 & & \oplus \displaystyle\bigoplus_{\begin{array}{c}
                                    \scriptstyle q'' \in Q\\[-2pt]
                                    \scriptstyle t = bv, b \in \Delta
                                    \end{array}} &
      \wei(q, \epsilon, b, q'') \otimes \weight_T(q'', s, v, q')\\ 
      & & \oplus \displaystyle\bigoplus_{\begin{array}{c}
                                         \scriptstyle q'' \in Q\\[-2pt]
                                         \scriptstyle s = au, a \in \Sigma\\[-2pt]
                                         \scriptstyle t = bv, b \in \Delta
                                         \end{array}} &
      \wei(q, a, b, q'') \otimes \weight_T(q'', u, v, q').\\ 
\end{array}
\]
We recall that, by convention, an empty sum with $\oplus$ is $\zero$. 
%
Since $\zero$ is absorbing for~$\otimes$ in~$\Semiring$,
one term $\wei(q, a, b, q'')$ equal to $\zero$ in the above expression 
will be ignored in the sum, meaning that there is no possible transition
from state $q$ into state $q'$ while reading $a$ and writing $b$ 
(including the case $a = \epsilon$ and $a = \epsilon$).
%These functions $\phi$ act as guards for the transducer's transitions, 
%preventing a transition when they return the absorbing $\zero$ of $\Semiring$.
This is analogous to a transition's guard not satisfied by $\<a, b>$ in 
the case of symbolic transducers.

The above expression of $\weight_T$ can be seen as a stateful definition of 
an edit-distance between a word $s \in \Sigma^*$ and a word $t \in \Delta^*$,
see also~\cite{Mohri03ijfcs}.
Intuitively, 
$\wei(q, a, \epsilon, q'')$, in the second line, is the cost of 
the deletion of a symbol $a \in \Sigma$ in $s$, 
$\wei(q, \epsilon, b, q'')$, in the third line, is the cost 
of the deletion of insertion of  $b \in \Delta$ in $s$, 
$\wei(q, a, b, q'')$, in the last line, is the cost 
of the substitution of  $a \in \Sigma$ by $b \in \Delta$.
The cost of a sequence of such operations transforming $s$ into $t$, 
is the product, with $\otimes$, of the individual costs of the operations involved;
And the distance between $s$ and $t$ is the sum, with $\oplus$,
of all such product of costs.
